\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-03-29 paper.qmd
still working on background

<!-- - Connection between NLDR and MDS-->
Historically, $k-D$ representations of $p-D$ data have been computed using multidimensional Scaling (MDS), which includes principal components analysis as a special case.  The $k-D$ representation can be considered to be a layout of points in $k-D$ produced by an embedding procedure that maps the data from $p-D$. In MDS, the $k-D$ layout is constructed by minimizing a stress function that differences distances between points in $p-D$ with potential distances between points in $k-D$. Various formulations of the stress function result in non-metric scaling [@article62] and isomap [@article63]. Challenges in working with high-dimensional data, including visualization, are outlined in @Johnstone2009.

@@

- Tours
- Model-in-the-data-space: how can we represent the model, eg plane for PCA, grid of values for classification boundaries.

>>>>

<!-- - Connection between NLDR and MDS-->
Historically, $k-D$ representations of $p-D$ data have been computed using multidimensional Scaling (MDS), which includes principal components analysis (PCA) as a special case.  The $k-D$ representation can be considered to be a layout of points in $k-D$ produced by an embedding procedure that maps the data from $p-D$. In MDS, the $k-D$ layout is constructed by minimizing a stress function that differences distances between points in $p-D$ with potential distances between points in $k-D$. Various formulations of the stress function result in non-metric scaling [@article62] and isomap [@article63]. Challenges in working with high-dimensional data, including visualization, are outlined in @Johnstone2009.

@@

The array of layouts in @fig-NLDR-variety illustrate what can emerge from the choices of method and parameters, and the random seed that initiates the computation. Key structures interpreted from these views suggest:

- highly **separated clusters** (a, b, e, g, h) with the number ranging from 3-6, 
- **stringy branches** (f), and 
- **barely separated clusters** (c, d) which apparently **contradicts** the other representations.

It happens because these methods and parameter choices provide different lenses on the interpoint distances in the data.

The alternative approach to visualizing the high-dimensional data is to use linear projections. PCA is the classical approach, resulting in a set of new variables which are linear combinations of the original variables. Tours, defined by @Asimov1985, broaden the scope by providing movies of linear projections, that provide views the data from all directions. @lee2021review provides an review of the main developments in tours. There are many tour algorithms implemented, with many available in the R package `tourr` [@article61], and versions enabling better interactivity in `langevitour` [@article09] and `detourr` [@article22]. Linear projections are a safe way to view high-dimensional data, because they do not warp the space, so they are more faithful representations of the structure. 
However, linear projections can be cluttered, and global patterns can obscure local structure. The simple activity of projecting data from $p-D$ suffers from piling [@laa2022], where data concentrates in the center of projections. NLDR is designed to escape these issues, to exaggerate structure so that it can be observed. But as a result NLDR can hallucinate wildly, to suggest patterns that are not actually present in the data.

The solution is to use the tour to examine how the NLDR is warping the space.

- Model-in-the-data-space: how can we represent the model, eg plane for PCA, grid of values for classification boundaries.

### 2024-03-29 paper.tex
still working on background

Historically, \(k-D\) representations of \(p-D\) data have been computed
using multidimensional Scaling (MDS), which includes principal
components analysis as a special case. The \(k-D\) representation can be
considered to be a layout of points in \(k-D\) produced by an embedding
procedure that maps the data from \(p-D\). In MDS, the \(k-D\) layout is
constructed by minimizing a stress function that differences distances
between points in \(p-D\) with potential distances between points in
\(k-D\). Various formulations of the stress function result in
non-metric scaling \citep{article62} and isomap \citep{article63}.
Challenges in working with high-dimensional data, including
visualization, are outlined in \citet{Johnstone2009}.

@@

\begin{itemize}
\tightlist
\item
  Tours
\item
  Model-in-the-data-space: how can we represent the model, eg plane for
  PCA, grid of values for classification boundaries.
\end{itemize}

>>>>

Historically, \(k-D\) representations of \(p-D\) data have been computed
using multidimensional Scaling (MDS), which includes principal
components analysis (PCA) as a special case. The \(k-D\) representation
can be considered to be a layout of points in \(k-D\) produced by an
embedding procedure that maps the data from \(p-D\). In MDS, the \(k-D\)
layout is constructed by minimizing a stress function that differences
distances between points in \(p-D\) with potential distances between
points in \(k-D\). Various formulations of the stress function result in
non-metric scaling \citep{article62} and isomap \citep{article63}.
Challenges in working with high-dimensional data, including
visualization, are outlined in \citet{Johnstone2009}.

@@

The array of layouts in Figure~\ref{fig-NLDR-variety} illustrate what
can emerge from the choices of method and parameters, and the random
seed that initiates the computation. Key structures interpreted from
these views suggest:

\begin{itemize}
\tightlist
\item
  highly \textbf{separated clusters} (a, b, e, g, h) with the number
  ranging from 3-6,
\item
  \textbf{stringy branches} (f), and
\item
  \textbf{barely separated clusters} (c, d) which apparently
  \textbf{contradicts} the other representations.
\end{itemize}

It happens because these methods and parameter choices provide different
lenses on the interpoint distances in the data.

The alternative approach to visualizing the high-dimensional data is to
use linear projections. PCA is the classical approach, resulting in a
set of new variables which are linear combinations of the original
variables. Tours, defined by \citet{Asimov1985}, broaden the scope by
providing movies of linear projections, that provide views the data from
all directions. \citet{lee2021review} provides an review of the main
developments in tours. There are many tour algorithms implemented, with
many available in the R package \texttt{tourr} \citep{article61}, and
versions enabling better interactivity in \texttt{langevitour}
\citep{article09} and \texttt{detourr} \citep{article22}. Linear
projections are a safe way to view high-dimensional data, because they
do not warp the space, so they are more faithful representations of the
structure. However, linear projections can be cluttered, and global
patterns can obscure local structure. The simple activity of projecting
data from \(p-D\) suffers from piling \citep{laa2022}, where data
concentrates in the center of projections. NLDR is designed to escape
these issues, to exaggerate structure so that it can be observed. But as
a result NLDR can hallucinate wildly, to suggest patterns that are not
actually present in the data.

The solution is to use the tour to examine how the NLDR is warping the
space.

\begin{itemize}
\tightlist
\item
  Model-in-the-data-space: how can we represent the model, eg plane for
  PCA, grid of values for classification boundaries.
\end{itemize}
